---
title: "Week 3 Assignment"
author: "Isaias soto"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##### \#1
*Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"*

```{r} 
library(tidyverse)
library(RCurl)
x <- getURL("https://raw.githubusercontent.com/isaias-soto/CUNY_DAT607/08e7bcca8ddd127ac68612a511c7d05f8d16261e/majors-list.csv") 
major_list <- read.csv(text = x)
major_list[146,]
major_list <- major_list[-146,] # remove missing value

```

```{r}
str_view(major_list$Major,"DATA|STATISTICS")
major_list[c(44,52,59),]
```


##### \#2 
*Write code that transforms the data below:*

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

*Into a format like this:*

*c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")*



The following code sets up the problem as a character vector with 14 elements:
```{r}
fruits <- c("bell pepper", "bilberry", "blackberry", "blood orange", 
            "blueberry", "cantaloupe", "chili pepper", "cloudberry", 
            "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")
fruits
```

Next, it seems we are to shrink this into a one-element vector. This can be achieved by the following:
```{r}
str_view(str_flatten(fruits, ", "))
```



*The two exercises below are taken from R for Data Science, 14.3.5.1 in the on-line version:*

##### \#3 

*Describe, in words, what these expressions will match:*

|Header|Description|

(.)\1\1
"(.)(.)\\\\2\\\\1"
(..)\1
"(.).\\\\1.\\\\1"
"(.)(.)(.).*\\\\3\\\\2\\\\1"


| Exercise                                	| Description                                                                                                                                                                                                                	|
|-----------------------------------------	|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------	|
| (.)\1\1                                 	| This would match words that have the same letter repeated 3 times, for example, "bbb".                                                                                                                                     	|
| "(.)(.)\\\\2\\\\1"                      	| This would match words where the second letter is repeated and this first letter is the start and end, for example, "abba".                                                                                                	|
| (..)\1                                  	| This would match words with any two letters with the same two letters repeated, for example, "cdcd".                                                                                                                       	|
| "(.).\\\\1.\\\\1"                       	| This would match words that begin with any letter, followed by a different letter, followed by the first letter repeated, <br>followed by any letter, and finally with followed by the first letter. For example, "abaca". 	|
| "(.)(.)(.).*\\\\3\\\\2\\\\1"            	| This matches words where there are 3 different letters, optionally followed by zero to any number of characters, and end with <br>the first three letters in reverse order. For example, "abcdcba" or "abccba".            	|

##### \#4 
*Construct regular expressions to match words that:*

*Start and end with the same character.*
*Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)*
*Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)*


| Exercise                                                                                  	| regex                   	|
|-------------------------------------------------------------------------------------------	|-------------------------	|
| Start and end with the same character.                                                    	| ^(.).*\\1$              	|
| Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)          	| ([A-Za-z][A-Za-z]).*\\1 	|
| Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.) 	| [A-Za-z].\*\\1.\*\\1    	|


****